Structure Learning for $\VAR(p)$ Models

### 2019-09-29 presentations/student_dynamical_systems_waterloo_2019/Abstract-Template.tex

\documentclass{statsoc}
% \usepackage{mathtime}
\usepackage{ccfonts,eulervm}
\usepackage{tabularx}
\usepackage{xr}
\externaldocument{supplementary_material}
% ---------------------------------------------------------------------------------------------------------------------------------
% The statsoc style file is broken?
% I have copied this code straight from stack overflow
% https://tex.stackexchange.com/questions/351768/problem-with-statsoc-class-and-pdflatex
\usepackage{geometry}

\geometry{
  textwidth=33pc,
  textheight=\dimexpr48\baselineskip+\topskip\relax,
  marginparsep=11pt,
  marginparwidth=107pt,
  footnotesep=6.65pt,
  headheight=9pt,
  headsep=9pt,
  footskip=30pt,
}
% ---------------------------------------------------------------------------------------------------------------------------------

% \usepackage{natbib}
\bibliographystyle{chicago}

%% bibliography stuff -- this needs come before the preamble inclusion
\usepackage{enumitem}
\usepackage{etex,etoolbox}
\usepackage{hyperref}
\usepackage{fullpage}
\setlength{\marginparwidth}{2cm}
\usepackage{todonotes}
\usepackage{multirow}
\usepackage{booktabs}

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\ngc{\overset{\text{GC}}{\nrightarrow}}  % Negated Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Pairwise Granger causality arrow
\def\npwgc{\overset{\text{PW}}{\nrightarrow}}  % Negated pairwise Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger causality graph
\def\gcge{\mathcal{E}}  % Graph edges
\def\VAR{\mathsf{VAR}}  % VAR(p) model
\def\MA{\mathsf{MA}}  % MA(p) model
\def\B{\mathsf{B}}  % Filter B
\def\wtB{\widetilde{\B}}  % General filter B
\def\A{\mathsf{A}}  % Filter A
\def\H{\mathcal{H}}  % Hilbert space

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{pa(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node
\newcommand{\ancn}[2]{\mathcal{A}_{#1}(#2)}  % nth ancestors of a node
\newcommand{\gpn}[2]{gp_{#1}(#2)}  % nth generation grandparents
\newcommand{\wtalpha}[2]{\widetilde{\alpha}(#1, #2)}  % Some notation for lem:pwgc_anc
\newcommand{\dist}[2]{\mathsf{d}(#1, #2)}  % Distance between things
\newcommand{\gcgpath}[2]{#1 \rightarrow \cdots \rightarrow #2}  % A shorter path command

% \usepackage{fullpage}
\usepackage{framed}

% Figures
\usepackage{graphicx}
\usepackage{caption}  % This is not recommended?
\usepackage{subcaption}
% \usepackage{wrapfig}
% \usepackage{svg}

% Math packages, theorem definitions and numbering
\usepackage{amsmath}
\usepackage{amssymb}
% \usepackage{amsthm}
% \usepackage{mathrsfs} % Fancy scripted font
\usepackage{bm}  % Bold math

% Misc packages
\usepackage[linesnumbered, ruled, vlined]{algorithm2e}
\usepackage{algorithm2e} %{algorithm} environment
\usepackage{soul}  % \hl highlighting
\usepackage{color}
\usepackage{mathtools}  % For my \ceil function

% Theorems (with italics)
% \theoremstyle{plain}  % Style definition removes italics
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}

% \theoremstyle{definition}
\newtheorem{remark}{Remark}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{assumption}{Assumption}

% Should I do this?  It makes cross references much easier,
% but is possibly extra complication.
% \counterwithin{theorem}{section}
% \counterwithin{corollary}{section}
% \counterwithin{proposition}{section}
% \counterwithin{lemma}{section}

% \counterwithin{remark}{section}
% \counterwithin{definition}{section}
% \counterwithin{example}{section}
% \counterwithin{assumptions}{section}

% keywords
\providecommand{\keywords}[1]{\textbf{\textit{Keywords---}} #1}

% General
\def\defeq{\overset{\Delta}{=}}  % Equal with triangle
\def\cl{\mathsf{cl\ }}  % Closure
\newcommand{\sgn}[1]{\mathsf{sgn}(#1)}  % sign function

% Calculus
\def\d{\mathsf{d}}  % Differential operator

% Functions
\def\ln{\mathsf{ln\ }}  % Natural logarithm
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}  % Ceiling

% Probability
\def\H{\mathcal{H}}  % Hilbert space
\def\E{\mathbb{E}}  % Expectation
\def\Var{\text{Var}}  % Variance
\def\P{\mathbb{P}}  % Probability Measure
\def\F{\mathcal{F}}  % A sigma algebra
\def\sX{\mathcal{X}}  % Another sigma algebra
\def\KL{\mathbf{D}_{KL}}  % KL divergence
\def\bF{\mathbf{F}}  % Whole F-meas space

% Standard sets
\def\Z{\mathbb{Z}}  % Set of integers
\def\R{\mathbb{R}}  % Set of real numbers
\def\C{\mathbb{C}}  % Set of complex numbers
\def\N{\mathbb{N}}  % Set of natural numbers
\def\ball{\mathbb{B}}  % Open ball
\def\clball{\overline{\ball}}  % Closed ball

% Linear algebra
\def\rk{\mathsf{rk }}  % The rank
\def\tr{\mathsf{tr }}  % The trace
\def\T{\mathsf{T}}  % Transpose notation
\def\c{\mathsf{c}}  % complement
\def\dg{\mathsf{dg }}   %  Diagonal vector of a matrix
\def\Dg{\mathsf{Dg }}   %  Diagonal matrix from a vector
\def\ind{\mathbf{1}}  % Ones vector or indicator
\def\matvec{\textbf{vec}}  % Vector operator
\def\<{\langle}  % < Inner product
\def\>{\rangle}  % > Inner product
\newcommand{\inner}[2]{\langle #1, #2 \rangle}  % Inner product
\newcommand{\innerT}[2]{#1^\T #2}  % Inner product for finite vectors

\graphicspath{{../figures/}}

\title{Graph Topological Aspects of Granger Causal Network Learning}
\author[Author 1 {\it et al.}]{R. J. Kinnear}
\address{
  University of Waterloo,
  Waterloo,
  Canada.}
\email{Ryan@Kinnear.ca}

\author{R. R. Mazumdar}
\address{
  University of Waterloo,
  Waterloo,
  Canada.}

\begin{document}

\abstract{We study Granger causality in the context of wide-sense
  stationary time series, where our focus is on the topological
  aspects of the underlying causality graph.  We establish sufficient
  conditions (in particular, we develop the notion of a ``strongly
  causal'' graph topology) under which the true causality graph can be
  recovered via pairwise causality testing alone, and provide examples
  from the gene regulatory network literature suggesting that our
  concept of a strongly causal graph may be applicable to this field.
  We implement and detail finite-sample heuristics derived from our
  theory, and establish through simulation the efficiency gains (both
  statistical and computational) which can be obtained (in comparison
  to LASSO-type algorithms) when structural assumptions are met.}

\keywords{causality graph, Granger causality, network learning, time
  series, vector autoregression, LASSO}

\paragraph{Acknowledgement}

We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC), [funding reference number 518418-2018].  Cette recherche a été financée par le Conseil de recherches en sciences naturelles et en génie du Canada (CRSNG), [numéro de référence 518418-2018].

\section{Introduction and Review}
\label{sec:introduction}
In this paper we study the notion of Granger causality
\cite{granger1969investigating} \cite{Granger1980329} as a means of
uncovering an underlying causal structure in multivariate time series.
Though the underlying causality graph cannot be observed directly,
it's presence is inferred as a latent structure among observed
time series data.  This notion is leveraged in a variety of
applications e.g. in Neuroscience as a means of recovering
interactions amongst brain regions \cite{bressler2011wiener},
\cite{anna_paper2008}, \cite{david2008identifying}; in the study of
the dependence and connectedness of financial institutions
\cite{NBERw16223}; gene expression networks \cite{Fujita2007},
\cite{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data},
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery},
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty};
and power system design \cite{Misyrlis2016450}, \cite{yuan2014root}.

Granger causality can generally be formulated by searching for the
``best'' graph structure consistent with observed data, which is in
general an extremely challenging problem (i.e. it may be framed as a
best subset selection problem, see, \cite{hastie_bss_comp} for recent
improvements in BSS methods), moreover, the comparison of quality
between different structures, and hence the notion of ``best'' needs
qualification.  In applications where we are interested merely in
minimizing the mean squared error of a linear one-step-ahead
predictor, then we will be satisfied with an entirely dense graph of
connections, since each edge can only serve to reduce estimation
error.  However, since the number of edges scales quadratically in $n$
(the number of nodes) it becomes imperative to infer a sparse
causality graph for large systems, both to avoid overfitting observed
data, as well as to aid the interpretability of the results.

A fairly early approach to the problem in the context of large systems
is provided by \cite{bach2004learning}, where the authors apply a
local search heuristic to the Whittle likelihood with an AIC
penalization.  The local search heuristic is a common approach to
combinatorial optimization due to it's simplicity, but is liable to
get stuck in shallow local minima.

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys fairly strong theoretical guarantees
\cite{wainwright2009sharp}, extending largely to the case of
stationary time series data with a sufficiently fast rate of
dependence decay \cite{basu2015} \cite{wong2016lasso}
\cite{autoregressive_process_modelling_via_the_lasso_procedure}, and
variations on the LASSO have been applied in a number of different
time series contexts as well as Granger causality
\cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery}.
One of the key improvements to the original LASSO algorithm is the
adaptive (i.e. weighted) ``adaLASSO'' \cite{adaptive_lasso_zou2006},
for which oracle results (i.e. asymptotic support recovery) are
established under less restrictive conditions than for the vanilla
LASSO.

% We study a novel perspective on the problem by establishing conditions
% under which simple \textit{pairwise} Granger causality tests are
% sufficient to recover the entire graph.  Without particular
% topological assumptions, pairwise testing naturally results in a large
% number of false positives \cite{tam2013gene_pwgc}.

In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger causality graph.  The focus of this paper is to shed light on
some of these topological questions, in particular, we study a
particularly simple notion of graph topology which we term
``strongly causal'' and show that stationary times series whose
underlying causality graph has this structure satisfy natural
intuitive notions of ``information flow'' through the graph.
Moreover, we show that such graphs are perfectly recoverable with only
\textit{pairwise} Granger causality tests, which would otherwise
suffer from serious confounding problems (see
e.g. \cite{tam2013gene_pwgc} for earlier work on pairwise testing).
Aside from being an interesting theoretical perspective, prior
assumptions about the underlying graph (similarly to sparsity
assumptions) can greatly improve upon the statistical power of
causality graph recovery algorithms when the assumptions are met.

Detailed study of Granger causality for star structured graphs has
been carried out in \cite{jozsa2018relationship}.  See as well
\cite{barnett2015granger}, \cite{jozsa2018_phd_thesis} for state space
formulations.

In the case of gene expression networks, we show examples from the
literature which suggest our concept of a ``strongly causal graph''
topology may have application in this field (see Section
\ref{sec:strongly_causal_graphs}).

@@

\section{Graph Topological Aspects of Granger causality}
\label{sec:theory}
\subsection{Formal Setting}
Consider the space $L_2(\Omega)$, the usual Hilbert space of finite
variance random variables over a probability space
$(\Omega, \mathcal{F}, \mathbb{P})$ having inner product
$\inner{x}{y} = \E[xy]$.  We will work with a discrete time and
wide-sense stationary (WSS) $n$-dimensional vector valued process
$x(t)$ (with $t \in \Z$) where the $n$ elements take values in $L_2$.
We suppose that $x(t)$ has zero mean, $\E x(t) = 0$, and has
absolutely summable matrix valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$, and an absolutely continuous
spectral density.

We will also work frequently with the spaces spanned by the values of
such a process

\begin{equation}
  \label{eq:hilbert_space_defn}
  \begin{aligned}
    \H_t^{(x)} &= \cl \{\sum_{\tau = 0}^p a_\tau^\T x(t - \tau)\ |\ a_\tau \in \R^n, p \in \N\} \subseteq L_2(\Omega)\\
    H_t^{(x)} &= \{a x(t)\ |\ a \in \R\} \subseteq L_2(\Omega),
  \end{aligned}
\end{equation}

where the closure is naturally in mean-square.  We will often omit the
superscript $x$ which should be clear from context.  Evidently these
spaces are separable, and as closed subspaces of a Hilbert space they
are themselves Hilbert.  We will denote the spaces generated in
analogous ways by particular components of $x$ as e.g.
$\H_t^{(i, j)}$, $\H_t^{i}$ or by all but a particular component as
$\H_t^{(-j)}$.

As a consequence of the Wold decomposition theorem (see
e.g. \cite{lindquist}), every WSS sequence has the moving average
$\MA(\infty)$ representation

\begin{equation}
  \label{eqn:wold}
  x(t) = c(t) + \sum_{\tau = 0}^\infty A(\tau) v(t - \tau),
\end{equation}

where $c(t)$ is a purely deterministic sequence, $v(t)$ is an
uncorrelated sequence and $A(0) = I$.  We will assume that $c(t) = 0$.
Given our setup, this representation can be inverted to yield the
$\VAR(\infty)$ form

\begin{equation}
  \label{eqn:ar_representation}
  x(t) = \sum_{\tau = 1}^\infty B(\tau) x(t - \tau) + v(t).
\end{equation}

The Equations \eqref{eqn:wold}, \eqref{eqn:ar_representation} can be
represented as $x(t) = \A(z)v(t) = \B(z)x(t) + v(t)$ via the action
(convolution) of the operators (LTI filters)
$$\A(z) \defeq \sum_{\tau = 0}^\infty A(\tau)z^{-\tau}$$ and
$$\B(z) \defeq \sum_{\tau = 1}^\infty B(\tau)z^{-\tau},$$ where the
operator $z^{-1}$ is the back shift operator acting on
$\ell_2^n(\Omega, \mathcal{F}, \mathbb{P})$, that is:

\begin{equation}
  \label{eqn:filter_action}
  \B_{ij}(z)x_j(t) \defeq \sum_{\tau = 1}^\infty B_{ij}(\tau)x_j(t - \tau).
\end{equation}

Finally, we have the inversion formula

\begin{equation}
  \label{eqn:lsi_inversion}
  \A(z) = (I - \B(z))^{-1} = \sum_{k = 0}^\infty \B(z)^k.
\end{equation}

The aforementioned assumptions are quite weak.  The strongest
assumption we require is finally that $\Sigma_v$ is a diagonal
positive-definite matrix, which is referred to as a lack of
instantaneous feedback in $x(t)$.  We formally state our setup as a
definition, which is the setup for the remainder of the paper:

\begin{definition}[Basic Setup]
  \label{def:basic_setup}
  The process $x(t)$ is an $n$ dimensional wide sense stationary
  process having invertible $\VAR(\infty)$ representation
  \eqref{eqn:ar_representation} where $v(t)$ is sequentially
  uncorrelated and has a diagonal positive-definite covariance matrix.
  The $\MA(\infty)$ representation of Equation \eqref{eqn:wold} has
  $c(t) = 0$ and $A(0) = I$.
\end{definition}

\subsection{Granger Causality}

\begin{definition}[Granger Causality]
  \label{def:granger_causality}
  For the WSS series $x(t)$ satisfying the assumptions of Definition
  \ref{def:basic_setup} we will say that component $x_j$
  \textit{Granger-Causes} (GC) component $x_i$ (with respect to $x$)
  and write $x_j \gc x_i$ if

\begin{equation}
  \linEerr{x_i(t)}{\H_{t - 1}} < \linEerr{x_i(t)}{\H^{(-j)}_{t - 1}},
\end{equation}

where $\xi[x \ |\ \H] \defeq \E (x - \linE{x}{\H})^2$ is the mean
squared estimation error and $\linE{x}{\H} \defeq \text{proj}_{\H}(x)$
denotes the (unique) projection onto the Hilbert space $\H$.
\end{definition}

This notion captures the idea that the process $x_j$ provides
information about $x_i$ that is not available from elsewhere.  The
caveat ``with respect to $x$'' is important in that GC relations can
change when components are added to or removed from our collection $x$
of observations, e.g. new GC relations can arise if we remove the
observations of a common cause, and existing GC relations can
disappear if we observe a new mediating series. The notion is closely
related to the information theoretic measure of transfer entropy,
indeed, if the distribution of $v(t)$ is known to be Gaussian then
they are equivalent \cite{barnett2009granger}.

The notion of conditional orthogonality is the essence of
Granger causality, and enables us to obtain results for a fairly
general class of WSS processes, rather than simply $\VAR(p)$ models.

\begin{definition}[Conditional Orthogonality]
  \label{lem:conditional_orthogonality_equivalence}
  Consider three closed subspaces of a Hilbert space $\mathcal{A}$,
  $\mathcal{B}$, $\mathcal{X}$.  We say that $\mathcal{A}$ is
  conditionally orthogonal to $\mathcal{B}$ given $\mathcal{X}$
  and write $\mathcal{A} \perp \mathcal{B}\ |\ \mathcal{X}$ if

    \begin{equation*}
      \inner{a - \linE{a}{\mathcal{X}}}{b - \linE{b}{\mathcal{X}}} = 0\ \forall a \in \mathcal{A}, b \in \mathcal{B}.
    \end{equation*}

  An equivalent condition is that (see \cite{lindquist} Proposition 2.4.2)
  \begin{equation*}
    \linE{\beta}{\mathcal{A} \vee \mathcal{X}} = \linE{\beta}{\mathcal{X}}\ \forall \beta \in \mathcal{B}
  \end{equation*}
\end{definition}

\begin{theorem}[Granger Causality Equivalences]
  \label{thm:granger_causality_equivalences}
  The following are equivalent:

  \begin{enumerate}
    \item{$x_j \ngc x_i$}
    \item{$\forall \tau \in \N_+\ B_{ij}(\tau) = 0$ i.e. $\B_{ij}(z) = 0$}
    \item{$H_t^{i} \perp \H_{t - 1}^{(j)}\ |\ \H_{t - 1}^{(-j)}$}
    \item{$\linE{x_i(t)}{\H_{t - 1}^{(-j)}} = \linE{x_i(t)}{\H_{t - 1}}$}
  \end{enumerate}
\end{theorem}

\subsection{Granger Causality Graphs}
We establish some graph theoretic notation and terminology, collected
formally in definitions for the reader's convenient reference.

\begin{definition}[Graph Theory Review]
  A \textit{graph} $\gcg = (V, \gcge)$ is simply a
  tuple of sets respectively called \textit{nodes} and \textit{edges}.
  Throughout this paper, we have in all cases
  $V = [n] \defeq \{1, 2, \ldots, n\}$.  We will also focus solely on
  \textit{directed} graphs, where the edges
  $\gcge \subseteq V \times V$ are \textit{ordered} pairs.

  A (directed) \textit{path} (of length $r$) from node $i$ to node
  $j$, denoted $\gcgpath{i}{j}$, is a sequence
  $a_0, a_1, \ldots, a_{r - 1}, a_r$ with $a_0 = i$ and $a_r = j$ such
  that $\forall\ 0 \le k \le r\ (a_k, a_{k + 1}) \in \gcge$, and where
  $(a_k, a_{k - 1})$ are \textit{distinct} for $0 \le k < r$.

  A \textit{cycle} is a path of length $2$ or more between a node and
  itself.  An edge between a node and itself $(i, i) \in \gcge$ (which
  we do not consider to be a cycle) is referred to as a \textit{loop}.

  A graph $\gcg$ is a \textit{directed acyclic graph} (DAG) if it is a
  directed graph and does not contain any cycles.
\end{definition}

\begin{definition}[Parents, Grandparents, Ancestors]
  A node $j$ is a \textit{parent} of node $i$ if $(j, i) \in \gcge$.
  The set of all $i$'s parents will be denoted $\pa{i}$, and we
  explicitly exclude loops as a special case, that is,
  $i \not\in \pa{i}$ even if $(i, i) \in \gcge$.

  The set of level $\ell$ \textit{grandparents} of node $i$, denoted
  $\gpn{\ell}{i}$, is the set such that $j \in \gpn{\ell}{i}$ if and
  only if there is a \textit{directed path} of length $\ell$ in $\gcg$
  from $j$ to $i$.  Clearly, $\pa{i} = \gpn{1}{i}$.

  Finally, the set of \textit{level $\ell$ ancestors} of $i$:
  $\ancn{\ell}{i} = \bigcup_{\lambda \le \ell}\gpn{\lambda}{i}$ is the
  set such that $j \in \ancn{\ell}{i}$ if and only if there is a
  directed path of length $\ell$ \textit{or less} in $\gcg$ from $j$
  to $i$.  The set of \textit{all ancestors} of $i$
  (i.e. $\ancn{n}{i}$) is denoted simply $\anc{i}$.

  Recall that we do not allow a node to be it's own parent, although
  unless $\gcg$ is a DAG, a node \textit{can} be it's own ancestor.
  We will occasionally need to explicitly exclude $i$ from $\anc{i}$,
  in which case we will write $\anc{i}\setminus \{i\}$.
\end{definition}

Our principal object of study will be a graph determined by
Granger causality relations as follows.

\begin{definition}[Causality graph]
  We define the Granger causality graph $\gcg = ([n], \gcge)$ to be the directed
  graph formed on $n$ vertices where an edge $(j, i) \in \gcge$ if and
  only if $x_j$ Granger-causes $x_i$ (with respect to $x$).  That is,
  $$(j, i) \in \gcge \iff j \in \pa{i} \iff x_j \gc x_i.$$
\end{definition}

The edges of the Granger causality graph $\gcg$ can be given a general
notion of ``weight'' by associating an edge $(j, i)$ with the
\textit{strictly causal} LTI filter $\B_{ij}(z)$ (see Equation
\eqref{eqn:filter_action}).  Thence, the matrix $\B(z)$ is analogous
to a \textit{weighted adjacency matrix}\footnote{\footnotesize We are using the
  convention that $\B_{ij}(z)$ is a filter with input $x_j$ and output
  $x_i$ so as to write the action of the system as $\B(z)x(t)$ with
  $x(t)$ as a column vector.  This competes with the usual convention
  for adjacency matrices where $A_{ij} = 1$ if there is an edge
  $(i, j)$.  In our case, the sparsity pattern of $\B_{ij}$ is the
  \textit{transposed} conventional adjacency matrix.} for the graph $\gcg$.  And,
in the same way that the $k^{\text{th}}$ power of an adjacency matrix
counts the number of paths of length $k$ between nodes,
$(\B(z)^k)_{ij}$ is a filter isolating the ``action'' of $j$ on $i$ at
a time lag of $k$ steps, this is exemplified in the inversion formula
\eqref{eqn:lsi_inversion}.

From the $\VAR$ representation of $x(t)$ there is clearly a tight
relationship between each node and it's parent nodes, the relationship
is quantified through the sparsity pattern of $B(z)$.  Similarly, the
following proposition is analogous to the definition of feedback free
processes of \cite{caines1975feedback} and provides an interpretation
of the sparsity pattern of $A(z)$ (from the MA representation of
$x(t)$) in terms of the causality graph $\gcg$.

\begin{proposition}[Ancestor Expansion]
  \label{prop:parent_expanding}
  The component $x_i(t)$ of $x(t)$ can be represented in terms of it's
  parents in $\gcg$:

  \begin{equation}
    \label{eqn:parent_expansion}
    x_i(t) = v_i(t) + \B_{ii}(z)x_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t).
  \end{equation}

  Moreover, $x_i$ can be expanded in terms of it's ancestor's $v(t)$
  components only:

  \begin{equation}
    \label{eqn:ancestor_expansion}
    x_i(t) = \A_{ii}(z)v_i(t) + \sum_{\substack{k \in \anc{i} \\ k \ne i}}\A_{ik}(z)v_k(t),
  \end{equation}

  where $\A(z) = \sum_{\tau = 0}^\infty A(\tau)z^{-\tau}$ is the filter from
  the Wold decomposition representation of $x(t)$, Equation
  \eqref{eqn:wold}.
\end{proposition}

This statement is ultimately about the sparsity pattern in the Wold
decomposition matrices $A(\tau)$ since
$x_i(t) = \sum_{\tau = 0}^\infty \sum_{j = 1}^n A_{ij}(\tau)v_j(t -
\tau)$.  The proposition states that if $j \not \in \anc{i}$ then
$\A_{ij}(z) = 0$.

\subsection{Pairwise Granger Causality}
\label{sec:pwgc}
Recall that Granger causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed, completely
explains the connection from $x_j$ to $x_i$.  In this section we study
\textit{pairwise} Granger causality, and seek to understand when
knowledge of pairwise relations is sufficient to deduce the true fully
conditional relations of $\gcg$.

\begin{definition}[Pairwise Granger causality]
  We will say that $x_j$ pairwise Granger-causes $x_i$ and write
  $x_j \pwgc x_i$ if $x_j$ Granger-causes $x_i$ with respect only to
  $(x_i, x_j)$.
\end{definition}

This notion is of interest for a variety of reasons.  From a purely
conceptual standpoint, we will see how the notion can in some sense
capture the idea of ``flow of information'' in the underlying graph,
in the sense that if $j \in \anc{i}$ we expect that $j \pwgc i$.  It may
also be useful for reasoning about the conditions under which
\textit{unobserved} components of $x(t)$ may or may not interfere with
inference in the actually observed components.  Finally, motivated
from a practical standpoint to analyze causation in large systems,
practical estimation procedures based purely on pairwise causality
tests are of interest since the computation of such pairwise relations
is substantially easier.

The following propositions are essentially lemmas used for the proof
of the upcoming Proposition \ref{prop:ancestor_properties}, but
remain relevant for providing intuitive insight into the problems at
hand.

\begin{proposition}
  Consider distinct nodes $i, j$ in a Granger causality graph
  $\gcg$.  If

  \begin{enumerate}[label=(\alph*)]
    \item{$j \not\in \anc{i}$ and $i \not\in \anc{j}$}
    \item{$\anc{i}\cap\anc{j} = \emptyset$}
  \end{enumerate}

  then $\H_t^{(i)} \perp \H_t^{(j)}$, that is,
  $\forall s, \tau \in \Z_+\ \E[x_i(t - s)x_j(t - \tau)] = 0$.  Moreover,
  this means that $j \npwgc i$ and $\linE{x_j(t)}{\H_t^{(i)}} = 0$.
\end{proposition}

\begin{remark}
  It is possible for components of $x(t)$ to be correlated at some
  time lags without resulting in pairwise causality.  For instance,
  the conclusion $j \npwgc i$ of Proposition
  \ref{prop:separated_ancestor_uncorrelated} will still hold even if
  $i \in \anc{j}$, since $j$ cannot provide any information about $i$
  that is not available from observing $i$ itself.
\end{remark}

\begin{proposition}
  Consider distinct nodes $i, j$ in a Granger causality graph $\gcg$.
  If

  \begin{enumerate}[label=(\alph*)]
    \item{$j \not\in \anc{i}$}
    \item{$\anc{i}\cap\anc{j} = \emptyset$}
  \end{enumerate}

  then $j \npwgc i$.
\end{proposition}

The previous result can still be strengthened significantly; notice
that it is possible to have some $k \in \anc{i} \cap \anc{j}$ where still
$j \npwgc i$, an example is furnished by the three node graph
$k \rightarrow i \rightarrow j$ where clearly
$k \in \anc{i}\cap\anc{j}$ but $j \npwgc i$.  We must introduce the concept
of a \textit{confounding} variable, which effectively eliminates the
possibility presented in this example.

\begin{definition}[Confounder]
  A node $k$ will be referred to as a \textit{confounder} of nodes
  $i, j$ (neither of which are equal to $k$) if
  $k \in \anc{i} \cap \anc{j}$ and there exists a path
  $\gcgpath{k}{i}$ not containing $j$, and a path $\gcgpath{k}{j}$ not
  containing $i$. A simple example is furnished by the ``fork'' graph
  $i \leftarrow k \rightarrow j$.
\end{definition}

\begin{proposition}
  \label{prop:ancestor_properties}
  If in a Granger causality graph $\gcg$ where $j \pwgc i$ then
  $j \in \anc{i}$ or $\exists k \in \anc{i} \cap\anc{j}$ which is a
  confounder of $(i, j)$.
\end{proposition}

>>>>

%    DO NOT CHANGE THE SETTINGS BELOW
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[12pt]{article}
\usepackage{epsfig}
\usepackage{graphicx,float}
\usepackage{url}
\usepackage[hidelinks,colorlinks=true,bookmarks=true]{hyperref}
%*******************************************
\setlength{\textheight}{24cm}		
\setlength{\textwidth}{17.5cm}		
\setlength{\oddsidemargin}{-0.5cm}
\setlength{\topmargin}{-1.25cm}
%*******************************************
 \pagestyle{empty}

\begin{document}
\hspace{-0.70cm}\rule{17.5cm}{0.01 in} \\
\vspace{-0.4cm}
\begin{flushleft}
\Large \textbf{\noindent
Insert the Title of Your Talk Here}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{0.5cm}
\normalsize
 %%%%%%             NAMES OF AUTHORS         %%%%%%
 %%%%%% NAME OF SPEAKER SHOULD BE UNDERLINED %%%%%%%%%
\normalsize{
 \underline{Ryan J. Kinnear}$^1$, Ravi R. Mazumdar$^1$
} \\
\vspace{5mm}
\textit{\footnotesize
 %%%%%% AFFILIATION OF AUTHORS %%%%%%
$^1$ Department of Electrical and Computer Engineering, University of Waterloo, Waterloo, Ontario N2L 3G1, Canada\\
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
}
\end{flushleft}

We study Granger causality in the context of wide-sense stationary
time series, where our focus is on the topological aspects of the
underlying causality graph.  We establish sufficient conditions (in
particular, we develop the notion of a ``strongly causal'' graph
topology) under which the true causality graph can be recovered via
pairwise causality testing alone, and provide examples from the gene
regulatory network literature suggesting that our concept of a
strongly causal graph may be applicable to this field.  We implement
and detail finite-sample heuristics derived from our theory, and
establish through simulation the efficiency gains (both statistical
and computational) which can be obtained (in comparison to LASSO-type
algorithms) when structural assumptions are met.

In this paper we study the notion of Granger causality
\cite{granger1969investigating} \cite{Granger1980329} as a means of
uncovering an underlying causal structure in multivariate time series.
Though the underlying causality graph cannot be observed directly,
it's presence is inferred as a latent structure among observed
time series data.  This notion is leveraged in a variety of
applications e.g. in Neuroscience as a means of recovering
interactions amongst brain regions \cite{bressler2011wiener},
\cite{anna_paper2008}, \cite{david2008identifying}; in the study of
the dependence and connectedness of financial institutions
\cite{NBERw16223}; gene expression networks \cite{Fujita2007},
\cite{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data},
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery},
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty};
and power system design \cite{Misyrlis2016450}, \cite{yuan2014root}.

@@

In a general Granger causality graph, there are natural
\textit{necessary} conditions for a node $j$ to pairwise cause node
$i$, namely that $j$ is an ancestor to $i$ or that there is a third
confounding node $k$.  On the other hand, we will see simple examples
for why \textit{sufficient} conditions are more nuanced: if $j$ causes
$i$ conditionally on all other nodes in the graph, it need not be the
case that $j$ causes $i$ when the two are observed in isolation.  We
examine a special \textit{graph topological} assumption we term
``strongly causal'' wherein a natural converse \textit{does} hold,
satisfying an intuitive notion of ``causality flow'' through the
graph.  We highlight examples from the literature on gene networks
suggesting that assumptions of strongly causal graph structure may not
be out of place in some application areas.

As a consequence of the causality graph being strongly causal, we show
that it is possible to recover the full conditional causality graph
through pairwise testing alone -- that is, it isn't necessary to
perform any conditional causality tests.

% $\bullet$ Abstracts may contain equations: Equations are to be centered horizontally, with equation numbers aligned with the right margin. See equation (\ref{eq1})
% \begin{equation}
% e^{i\pi}+1=0.\label{eq1}
% \end{equation}

% $\bullet$ Abstracts must contain at least two  references. See   \cite{ASY,STE}.

% $\bullet$ Abstracts may contain figures. For example, Figure \ref{fig1}.

\begin{figure}[H]
\centering
\framebox[6cm]{\rule{0cm}{0.5cm} Abstracts may contain figures}
%\includegraphics[width =0.48\linewidth]{image} 
\caption{Figure one caption.}
\label{fig1}
\end{figure}

 Please send your source file  and .pdf file to Alyssa Novelia (anovelia@uwaterloo.ca) or Isam Al-Darabsah (ialdarabsah@uwaterloo.ca)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BIBLIOGRAPHY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{00}
\addcontentsline{toc}{chapter}{References}
\setlength{\itemsep}{-1mm}
\small
\bibitem{ASY} K.T. Alligood, T.D. Sauer and J. A. Yorke. \textit{Chaos: An Introduction to Dynamical Systems}, New York, Springer-Verlag (1997).

\bibitem{STE} L. Schimansky-Geier, A.V. Tolstopjatenko and W. Ebelin \textit{Noise induced transitions due to external additive noise}, Phys. Lett. A, \textbf{108}, 7, pp. 329-332 (1985).
\end{thebibliography}
\normalsize
\end{document}
